---
title: "Introduction to ginteff"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to ginteff}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment  = "#>",
    fig.width = 6, fig.height = 4
)
```

## What this package computes

In a model with one or more interactions, the *interaction effect* on
the response scale is rarely the same thing as the interaction *coefficient*.
Once you go through a non-linear link, the relevant quantity for inference
is the cross-partial of the predicted response. **ginteff** computes
that quantity -- for two- and three-way interactions, on the response
scale -- and gives delta-method standard errors that respect any
variance-covariance specification you pass through.

Two paths are available:

* `dydxs`     — the cross-partial derivative
                $\partial^k\,\mathrm{E}[y\mid x]/\partial x_1\cdots\partial x_k$,
                approximated numerically.
* `firstdiff` — the discrete cross-difference,
                $\Delta^k\,\mathrm{E}[y\mid x]/\Delta x_1\cdots\Delta x_k$,
                where each $\Delta x_j$ is a user-specified unit increase.

The package wraps `marginaleffects::avg_predictions()` to do the heavy
lifting, then assembles the signed linear combination by hand so that
arbitrary `vcov =` arguments (`"HC3"`, `~cluster`, sandwich matrices)
propagate cleanly to the final SE.

## Setup

```{r setup, message = FALSE, warning = FALSE}
library(ginteff)

dat <- read.csv(system.file("extdata", "simdata.csv", package = "ginteff"))

dat$female_f <- factor(dat$female, levels = c(0L, 1L),
                       labels = c("male", "female"))
dat$race_f   <- factor(dat$race,   levels = 1:3,
                       labels = c("white", "black", "other"))
dat$health_o <- ordered(dat$health, levels = 1:5)

nrow(dat); mean(dat$health_2l)
```

## Example 1 -- two factors

The Stata help opens with the discrete change of `female` × `race`
on the probability of poor health, holding the other covariates at
observed values. Here is the same call in R.

```{r ex1}
m1 <- glm(health_2l ~ female_f * race_f + age + height + weight,
          data = dat, family = binomial)

g1 <- ginteff(m1, dydxs = c("female_f", "race_f"), level = 0.90)
g1
```

The factor `race_f` has three levels (`white`, `black`, `other`) so
ginteff emits one row per non-base level. The numbers above match
Stata's `ginteff, dydxs(female race) level(90)` to all printed digits.

## Example 2 -- three continuous variables, partial derivative

```{r ex2}
m2 <- glm(health_2l ~ age * height * weight + female_f + race_f,
          data = dat, family = binomial)

ginteff(m2, dydxs = c("age", "height", "weight"))
```

## Example 3 -- the same effect via first difference

```{r ex3}
ginteff(m2, firstdiff = c("age", "height", "weight"),
        nunit = c(age = 1, height = 1, weight = 1))
```

The point estimate matches Example 2 (the partial derivative is the
limit of the first difference as the unit shrinks); the SE is slightly
different because the FD is an exact linear combination of predictions
while `dydxs` involves a numerical derivative. For three-way effects,
`firstdiff` gives the most accurate SE; see `?ginteff` for the
recommended default.

## Example 4 -- ordered logit

```{r ex4, eval = requireNamespace("MASS", quietly = TRUE)}
m4 <- MASS::polr(health_o ~ age * female_f, data = dat, Hess = TRUE)

ginteff(m4, dydxs = c("age", "female_f"),
        atdxs = list(age = "mean"),
        eqn   = 2L)            # outcome category 2
```

The `eqn` argument selects which outcome category's predicted
probability we are differentiating; for `polr` it can be a label
(`"2"`) or an integer position (`2L`).

## Robust standard errors

`vcov` is forwarded straight to `marginaleffects::avg_predictions()`,
so anything that package accepts works:

```{r robust, eval = requireNamespace("sandwich", quietly = TRUE)}
ginteff(m2, dydxs = c("age", "height", "weight"), vcov = "HC3")
```

Cluster-robust SEs:

```{r cluster, eval = FALSE}
ginteff(m2, dydxs = c("age", "height", "weight"), vcov = ~race)
```

A pre-computed vcov matrix:

```{r matvcov, eval = requireNamespace("sandwich", quietly = TRUE)}
V <- sandwich::vcovCL(m2, cluster = ~race)
ginteff(m2, dydxs = c("age", "height", "weight"), vcov = V)
```

## Per-observation effects

`obseff = TRUE` keeps the per-observation interaction effect (point
estimates only) so you can plot the heterogeneity behind the AIE:

```{r obseff}
g <- ginteff(m1, dydxs = c("female_f", "race_f"), obseff = TRUE,
             level = 0.90)
str(g$obseff)
```

If `ggplot2` is installed, `ginteffplot()` (or `plot()`) renders the
AIE on a horizontal axis with a capped-spike CI and (optionally) a
thin strip of the per-observation effects below -- the same layout as
Stata's `ginteffplot`:

```{r plot, eval = requireNamespace("ggplot2", quietly = TRUE)}
ginteffplot(g, output = 1L, obseff = TRUE)
```

To see the per-observation effects against a covariate (an R extension,
not in Stata's `ginteffplot`), pass `against = ` and the data:

```{r plot_against, eval = requireNamespace("ggplot2", quietly = TRUE)}
ginteffplot(g, output = 1L, obseff = TRUE,
            against = "age", data = dat)
```

## Tidy output for downstream use

```{r tidy}
as.data.frame(ginteff(m2, firstdiff = c("age", "height", "weight"),
                      nunit = c(age = 1, height = 1, weight = 1)))
```

## When to prefer `firstdiff` vs `dydxs`

* **two-way interactions**: either path is fine; the SEs match.
* **three-way interactions on a non-linear link**: prefer
  `firstdiff`. The numerical cross-partial in `dydxs` involves the
  finite-difference quotient `(2h)^{-3}`, which amplifies any noise in
  `marginaleffects`'s internally-computed prediction Jacobian; in
  testing the resulting SE comes out ~10 % lower than Stata's analytic
  computation. The `firstdiff` SE is exact regardless of `h`.

## Argument map vs Stata

| Stata                  | R                                                |
|------------------------|--------------------------------------------------|
| `dydxs(x1 x2)`         | `dydxs = c("x1", "x2")`                          |
| `firstdiff(x1 x2)`     | `firstdiff = c("x1", "x2")`                      |
| `nunit((1) x1 x2)`     | `nunit = c(x1 = 1, x2 = 1)`                      |
| `at((mean) x3)`        | `at = list(x3 = "mean")`                         |
| `atdxs((mean) x1)`     | `atdxs = list(x1 = "mean")`                      |
| `obseff(stub)`         | `obseff = TRUE`                                  |
| `level(95)`            | `level = 0.95`                                   |
| `intequation(#1)`      | `eqn = "outcome_label"` or integer index         |
| `vce(unconditional)`   | `vcov = ~cluster` / `"HC3"` etc.                 |
